---
title: "ST558 Project 2"
author: Dane Korver
date: Wednesday, November 6, 2024
format: html
editor: visual
---

```{r}
#| message: FALSE
#| echo: FALSE
#| warning: FALSE

# Load libraries
library(tidyverse)
# Read in data
nfl_plays <- read_csv("/Users/danekorver/Documents/Fall2024 ST558/scratch/NFL Play by Play 2009-2018 (v5).csv")
```

# Preparing for R Shiny app development

For this project, I decided to look at the [NFL play-by-play Data](https://www.kaggle.com/datasets/maxhorowitz/nflplaybyplay2009to2016/data) for the years 2009-2018 from kaggle.com. The data had 449,371 observations and 255 variables. Since each observation represents a play, let's look at a frequency distribution of the different play types between 2009-2018. From the output below, 41.54% of the plays between 2009-2018 were 'pass' plays and 29.53% were 'run' plays.

```{r}

nfl_plays %>% 
  count(play_type) %>%
  mutate(percent=n*100/sum(n))

```

If you watch football, you know that many games are decided by making field goals, which were 2.18% of all plays played between 2009-2018 and 83.93% of them were 'made'.

```{r}

nfl_plays %>% 
  filter(play_type=="field_goal") %>%
  count(field_goal_result) %>%
  mutate(percent=n*100/sum(n))

```

The average distance of the field goals 'made' between 2009-2018 was around 36 yards with a range of 18-64 yards. The record for the longest field goal was broken during this time frame in 2013 by Matt Prater of the Denver Broncos with a 64 yard field goal kick. Until that time, the record was 63 yards by Tom Dempsey of the New Orleans Saints in 1970. The current record is 66 yards held by Justin Tucker of the Baltimore Ravens which was kicked in the 2021 season. (Source: [ESPN](https://www.espn.com/nfl/story/_/id/40994229/what-longest-field-goals-nfl-history) and [Wikipedia](https://en.wikipedia.org/wiki/Field_goal))

```{r}

fg_made <- nfl_plays %>% 
  filter(play_type=="field_goal" & field_goal_result=="made")
summary(fg_made$kick_distance)

```
```{r}

ggplot(data=fg_made,aes(x=kick_distance)) + geom_bar(stat="count") + labs(title = "Histogram of 'Made' Field Goals", subtitle = "for the years 2009-2018", x = "Kick Distance (in yards)", y = "Count") + theme_minimal() + 
scale_x_continuous(breaks=seq(0,65,5)) +
scale_y_continuous(breaks=seq(0,400,50)) + 
theme(plot.title=element_text(hjust = 0.5),
        plot.subtitle=element_text(hjust = 0.5),
        axis.line = element_line(colour = "grey50"))

```